\section{Research Outline}

Hytale is a sandbox RPG whose identity is closely connected to modding, creator tools, and community-made content. In many games, modding begins after release, when players reverse-engineer systems, edit assets, or build unofficial tools to change the game. Hytale approaches this differently. It presents creation as part of the game's basic promise, describing itself as ``an adventure built for both creation and play.'' This makes Hytale useful for asking what happens when modding is no longer outside the game, but becomes part of how the game is imagined, developed, and shared.

For my report, I am especially interested in this question from the perspective of computer science. Hytale does not only invite players to create content visually, but also provides a development toolkit for programming plugins. This raises a technical question: what are the limits of what can be changed through a plugin? My own mod will therefore not mainly be an asset pack or a cosmetic extension. It will be a coded Hytale plugin that changes the rules of play and tests how much control the toolkit gives to modders.

Hytale mods are created through a mix of official and semi-official tools. Important tools include the Hytale Asset Editor, direct data asset editing, Blockbench for models and animations, and Java server plugins for gameplay changes. These tools suggest that Hytale modding is meant for different kinds of creators: artists, builders, server owners, and programmers. Modding is not limited to changing textures or adding small cosmetic details. It can also involve new systems, multiplayer rules, custom items, and changes to how the game behaves.

The main public distribution platform for Hytale mods is CurseForge, especially because public submissions, collections, and official contests are organized there. This means that mods are not only private experiments, but public artifacts that can be shared, downloaded, updated, and discussed by a wider community. At the same time, Hytale modding is still an evolving infrastructure, visible in the changing API, the unfinished tooling, and the way the developers continue to adjust what modders can access.

A key part of Hytale's modding strategy is that it is server-side first. Server hosts choose which mods are active, while players can join modded servers without manually installing external files. Modding therefore becomes part of the multiplayer structure of the game, not just an individual modification. For a plugin-based mod, this is especially relevant, because the server becomes the place where rules, events, and interactions can be rewritten.

Hytale modding involves several types of creators because its main formats are split between packs and plugins. Packs are focused on assets and content, such as blocks, mobs, items, textures, models, and cosmetics. Plugins are more code-based and can affect commands, events, server systems, minigames, and deeper gameplay rules. This creates a broad modding landscape that includes artists, programmers, designers, server hosts, toolmakers, documentation writers, and community organizers.

One useful example is Violet, the creator of \textit{Violet's Wardrobe}, a cosmetic mod that adds craftable customization options. The mod shows the importance of visual and cosmetic creation in Hytale, but Violet is also important as a modder because she later began working with the Hytale team on official cosmetics. This shows how modding can become a visible form of creative work and even a path into the official development process. In this case, modding is not only fan activity, but also a portfolio and a bridge between community creation and professional development.

The modding landscape has also changed because Hytale itself has had an unusual development history. The game was announced years ago, acquired by Riot Games, cancelled in 2025, and later revived after the original creators reacquired it. Reporting on the revival emphasizes that Hytale's return was connected to its original creator-focused vision and modding support from day one. Hytale is therefore not only a game that allows mods. It is a game whose public image and future are strongly tied to the promise of community creation.

Early Hytale modding also shows how unpredictable creator communities can be. Shortly after launch, players created experimental mods that ran Windows 95, Minecraft, and even Hytale itself inside Hytale. These are not necessarily typical gameplay mods, but they show that strong modding tools can turn a game into a platform for unexpected experiments.

Overall, Hytale is a useful case study because it shows modding becoming part of the development pipeline. Mods are created with official tools, shared through public platforms, used to build multiplayer spaces, and sometimes recognized by the developers themselves. My own project approaches this from the plugin side: instead of asking only what players can build, I want to ask what parts of the game's logic can be changed when the developers provide the programming interface.

\section{Critical Mod Proposal}

\subsection{Trial of the Goddess}

For my own critical mod, I plan to develop a Hytale plugin called \textit{Trial of the Goddess}. The mod will not focus mainly on new textures, models, or visual assets. Instead, it will use the Hytale development toolkit to interfere with the game logic itself. My goal is to test what kinds of rules, events, combat values, and player interactions can be changed through a plugin.

The starting point of the mod is Hytale's statues. At the moment, these statues mainly seem to function as mysterious worldbuilding objects. They create atmosphere and suggest lore, but they do not yet seem to have a strong gameplay role. In \textit{Trial of the Goddess}, a statue becomes the place where the player can begin a trial. The Goddess is not a full NPC, but a mythical presence connected to the statue: a being above the player who offers a challenge.

The goal of the trial is to retrieve a sacred flower and bring it back to the Goddess without dying. When the player starts the trial, they receive \textit{The Blade of Balance}. This is a powerful sword that can defeat enemies in one hit. However, while holding it, the player also only has one hit point and can be killed in one hit. The weapon therefore gives the player extreme power and extreme vulnerability at the same time.

Once the trial begins, the monster spawn rate in the area increases. Somewhere nearby, the sacred flower spawns. The player has to find it, pick it up, and return it to the statue. If the player succeeds, the trial is completed. If the player dies, the Blade of Balance is taken away, and the trial has to be restarted.

This makes the mod critical because it changes how players approach combat. Normally, becoming powerful in a game means becoming safer. \textit{Trial of the Goddess} inverts this logic: the player receives a mighty weapon, but that weapon also makes them fragile. Combat becomes less about slowly reducing enemy health and more about movement, timing, risk, and avoiding mistakes.

From a technical perspective, I want to test whether a Hytale plugin can react to player actions in real time, modify health and damage values, control item effects, change spawning conditions, and create a temporary rule system around a specific area or event. A successful first draft would already be a small working plugin where the player can start the trial at a statue, receive the Blade of Balance, search for the sacred flower, and either complete the trial or lose the blade after dying.

\subsection{Prototype Repository}

As part of the proposal, I also created a small first prototype of \textit{Trial of the Goddess}. The prototype is available here:

\begin{center}
    \url{https://github.com/ayanomaliy/GoddessTrial.git}
\end{center}

The repository contains my current experimental implementation of the plugin. It is not a finished version of the mod yet, but a proof of concept and a technical starting point. It gives me a concrete codebase to extend during the final report phase. Instructions for how to start and test the prototype can be found in the repository's README file.